Learning Action Models from State Observations

### 2018-01-26 doc/ijcai-18/observations-learning.tex
ijcai18 paper updated to replace figures
ijcai18 paper updated with friday meeting comments

Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016hierarchical,segovia2017generating}, we present a novel planning compilation approach for learning \strips\ action models. The output of the compilation is a classical planning task for which a solution plan is a sequence of actions that determine the precondition/effects of the model operator schemas (the learned model). \textbf{Learning action models as planning} leverages off-the-shelf planners and opens up a way towards the \emph{bootstrapping} of planning action models, enabling a planner to gradually learn and update its action model. The practicality of the compilation approach allow us to report results over a wide range of 15 IPC planning domains.

@@

This paper is organized as follows ...

@@

\subsubsection{Classical planning with conditional effects}
We use $F$ to denote the set of {\em fluents} (propositional variables) describing a state. A {\em literal} $l$ is a valuation of a fluent $f\in F$, i.e. either~$l=f$ or $l=\neg f$. A set of literals $L$ represents a partial assignment of values to fluents (WLOG we assume that $L$ does not assign conflicting values to any fluent). We use $\mathcal{L}(F)$ to denote the set of all literal sets on $F$, i.e.~all partial assignments of values to fluents.

@@

\section{Learning \strips\ action models}
Learning \strips\ action models from plans where every action in the plan is available as well as its corresponding {\em pre-} and {\em post-states}, is straightforward. In this case, \strips\ operator schemes are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Preconditions are derived lifting the minimal set of literals that appears in all the corresponding pre-states.

This paper addresses a more challenging learning task, where less input knowledge is available. The addressed learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions, the actual executed actions are unobserved. This learning task is defined as $\Lambda=\tup{\Xi,\Psi,\mathcal{O}}$, where:
\begin{itemize}
\item $\Xi$ is the set of empty operator schemes, wherein each $\xi\in\Xi$ is only composed of $head(\xi)$.
\item $\Psi$ is the set of predicates, that define the abstract state space of a given classical planning frame.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$.
\end{itemize}

A solution to $\Lambda$ is a set of operator schema $\Xi'$ compliant with the headers $\Xi$, the predicates $\Psi$, and the given sequence of state observations $\mathcal{O}$. A planning compilation is a suitable approach for addressing a $\Lambda$ learning task because a solution must not only determine the \strips\ action model $\Xi'$ but also, the {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$, that explains $\mathcal{O}$. Figure~\ref{fig:lexample} shows a $\Lambda$ task for learning a \strips\ action model in the blocksworld from the sequence of five state observations that corresponds to inverting a 2-blocks tower.

\begin{figure}
{\footnotesize\tt ;;;;;; Headers in $\Xi$}
\begin{footnotesize}
\begin{verbatim}
(pickup v1) (putdown v1)
(stack v1 v2} (unstack v1 v2)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;;;;; Predicates $\Psi$}
\begin{footnotesize}
\begin{verbatim}
(handempty) (holding ?o  - object)
(clear ?o - object) (ontable ?o - object)
(on ?o1 - object ?o2 - object)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;;;;; Observations $\mathcal{O}$}
\begin{footnotesize}
\begin{verbatim}
;;; observation #0
(clear block2) (on block2 block1)
(ontable block1) (handempty)

@@

Our approach for addressing the learning task is compiling $\Lambda$ into a classical planning task $P_{\Lambda}$ with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:

\begin{enumerate}
\item {\bf Programs the \strips\ action model $\Xi'$}. A solution plan starts with a {\em prefix} that, for each $\xi\in\Xi$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item {\bf Validates the \strips\ action model $\Xi'$ in $\mathcal{O}$}. The solution plan continues with a postfix that produces the given sequence of states $\tup{s_0,s_1,\ldots,s_{n}}$ using the programmed action model $\Xi'$.
\end{enumerate}

Given a learning task $\Lambda=\tup{\Xi,\Psi,\mathcal{O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ built instantiating the predicates $\Psi$ with the objects appearing in the input observations $\mathcal{O}$.
\item Fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$, that represent the programmed action model. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the \strips\ operator schema $\xi\in \Xi'$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) is represented by the pair of fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to {\em True}.
\item Fluent $mode_{prog}$ indicating whether the operator schemes are programmed or validated (already programmed) and fluents $\{test_i\}_{1\leq i\leq n}$, indicating the observation where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. Our compilation assumes that initially operator schemes are programmed with every possible precondition, no negative effect and no positive effect. Therefore fluents $pre_f(\xi)$, for every $f\in F_v(\xi)$, hold also at the initial state.
\item $G_{\Lambda}=\bigcup_{1\leq i\leq n}\{test_i\}$, indicates that the programmed action model is validated in all the input observations.
\item $A_{\Lambda}$ comprises three kinds of actions:
\begin{enumerate}
\item Actions for {\em programming} operator schema $\xi\in\Xi$:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\Xi$.

@@

\item Actions for {\bf adding} a {\em negative} or {\em positive} effect $f\in F_v(\xi)$ to the action schema $\xi\in\Xi$.

@@

\item Actions for {\em applying} an already programmed operator schema $\xi\in\Xi$ bound with the objects $\omega\subseteq\Omega^{ar(\xi)}$. Given that the operators headers are known, the variables $pars(\xi)$ are bound to the objects in $\omega$ appearing at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\}.
\end{align*}
\end{small}

\item Actions for {\em validating} an observation {\tt\small $1\leq i\leq n$}.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&s_i\cup\{test_j\}_{j\in 1\leq j<i}\\
&\cup\{\neg test_j\}_{j\in i\leq j\leq n}\cup \{\neg mode_{prog}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i\}.
\end{align*}
\end{small}
\end{enumerate}
\end{itemize}

\begin{figure}[hbt!]
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small Action for applying an already programmed schema $stack$ as encoded in PDDL (implications are coded as disjunctions).}
\label{fig:compilation}
\end{figure}

@@

\begin{lemma}
Soundness. Any classical plan $\pi$ that solves $P_{\Lambda}$ induces an action model $\Xi'$ that solves the $\Lambda$ learning task.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
The compilation forces that once the preconditions of an operator schema $\xi \in \Xi'$ are programmed, they cannot be altered. The same happens with the positive and negative effects. Furthermore, once operator schemes $\Xi'$ are programmed, they can only be applied because of the $mode_{prog}$ fluent. Finally, $P_{\Lambda}$ is only solvable if fluents $\{test_i\}$, {\small $1\leq i\leq n$} hold at the last reached state. These goals can only be achieved executing an applicable sequence of programmed operator schemes that reaches every state $s_i\in\mathcal{O}$, starting from $s_0$ and following the sequence {\small $1\leq i\leq n$} which means that the programmed action model $\Xi'$ complies with the provided observations and hence, solves $\Lambda$.
\end{small}
\end{proof}

\begin{lemma}
Completeness. Any \strips\ action model $\Xi'$ that solves a $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$ learning task, is computable solving the corresponding classical planning task $P_{\Lambda}$.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
The compilation does not discard any possible \strips\ action schema definable within $F_v$ that satisfy the state trajectory constraints given by the $\mathcal{O}$ sequence. By definition, $F_v(\xi)\subseteq F_\Lambda$ fully captures the full set of elements that can appear in a \strips\ action schema $\xi\in\Xi$ given its header and the set of predicates $\Psi$.
\end{small}
\end{proof}

The size of the compiled classical planning instances depends on the number of input examples. The larger the number of examples, the larger the compilation (a larger number of "test" fluents and "validate" actions), and also the more actions the solution plan requires to derive the action model.

\section{Evaluation of \strips\ action models}
We assess how well a \strips\ action model $\Xi$ explains a given sequence of observations $\mathcal{O}$ according to the amount of {\em edition} that is required by $\Xi$ to produce $\mathcal{O}$. In the extreme, if $\Xi$ perfectly explains $\mathcal{O}$, no model {\em edition} is necessary.

\subsection{Edition of \strips\ action models}
Let us define first the \emph{operations} allowed to edit a given \strips\ action model. With the aim of keeping tractable the branching factor of the planning instance that results from our compilation, we only define two {\em edit operations}:
\begin{itemize}
\item {\em Deletion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, such that $f\in F_v(\xi)$, is removed from the operator schema $\xi\in\Xi$.
\item {\em Insertion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, such that $f\in F_v(\xi)$, is added to the operator schema $\xi\in\Xi$.
\end{itemize}

We can now formalize an edit distance that quantifies how dissimilar two given \strips\ action models are. The distance is symmetric and meets the {\em metric axioms} provided that the two {\em \strips\ edit operations} have the same positive cost.

\begin{definition}
Let $\Xi$ and $\Xi'$ be two \strips\ action models, both built from the same set of possible elements $F_v$. The {\bf edit distance}, denoted as $\delta(\Xi,\Xi')$, is the minimum number of {\em edit operations} required to transform $\Xi$ into $\Xi'$.
\end{definition}

Since $F_v$ is a bound set, the maximum number of edits that can be introduced to a given action model defined within $F_v$ is bound as well. In more detail, for an operator schema $\xi\in\Xi$ the maximum number of edits that can be introduced to their precondition set is $|F_v(\xi)|$ while the max number of edits that can be introduced to the effects are two times $|F_v(\xi)|$.

\begin{definition}
The \textbf{maximum edit distance} of an \strips\ action model $\Xi$ built from the set of possible elements $F_v$ is $\delta(\Xi,*)=\sum_{\xi\in\Xi} 3|F_v(\xi)|$ and sets an upperbound on the distance from $\Xi$ to any other \strips\ action model definable within $F_v$.
\end{definition}

@@

\begin{definition}
Given an action model $\Xi$ built from the set of possible elements $F_v$ and a sequence of observations $\mathcal{O}$. The {\bf observation edit distance}, denoted by  $\delta(\Xi,\mathcal{O})$, is the minimal edit distance from $\Xi$ to any model $\Xi'$ such that: (1) $\Xi'$ is also definable within $F_v$ and (2), $\Xi'$ can produce a plan $\pi=\tup{a_1, \ldots, a_n}$ that induces $\mathcal{O}=\tup{s_0, s_1, \ldots, s_n}$, the given observations sequence; i.e., \[\delta(\Xi,\mathcal{O})=\min_{\forall \Xi' \rightarrow \mathcal{O}} \delta(\Xi,\Xi')\]
\end{definition}

Because its semantic nature, the {\em observation edit distance} is robust to learning episodes where actions are {\em reformulated} and still compliant with the learning inputs. For instance a learning episode can interchange the roles of operators with matching headers, or the roles of action parameters with matching types (e.g. the {\em blocksworld} operator {\small\tt stack} can be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa).

Furthermore, the {\em observation edit distance} enables the recognition of \strips\ action models. The idea, taken from {\em plan recognition as planning}~\cite{ramirez2009plan}, is to map distances into likelihoods. We can map the {\em observation edit distance} into a likelihood with the following expression $P(\mathcal{O}|\Xi)=1-\frac{\delta(\Xi,\mathcal{O})}{\delta(\Xi,*)}$. The larger the {\em observation edit distance} the lower this likelihood.

%Given a set of possible \strips\ models and a sequence of state observations, the {\em recognition of \strips\ models} is the task of computing the model with the highest probability according to the cited observations. According to the Bayes rule, the probability of an hypothesis $\mathcal{H}$ given the observations $\mathcal{O}$ can be computed with $P(\mathcal{H}|\mathcal{O})=\frac{P(\mathcal{O}|\mathcal{H})P(\mathcal{H})}{P(\mathcal{O})}$. In our scenario, the hypotheses are about the possible \strips\ action models that can be built within a given set of predicates $\Psi$ and a given a set of operator headers (in other words, given the $F_v(\xi)$ sets). With this regard, $P(\Xi|\mathcal{O})$, the probability distribution of the possible \strips\ models (within the $F_v(\xi)$ sets) given an observation sequence $\mathcal{O}$ could be computed by:
%\begin{enumerate}
%\item Computing the {\em observation edit distance} $\delta(\Xi,\mathcal{O})$ for every possible model $\Xi$.
%\item Applying the resulting distances to the above $P(\mathcal{O}|\Xi)$ formula to map these distances into likelihoods
%\item Applying the Bayes rule to obtain the normalized posterior probabilities, these probabilities must sum 1.
%\end{enumerate}

\subsection{Computing the observation edit distance}
Our compilation is extensible to compute the {\em observations edit distance}. A solution to the planning task resulting from the extended compilation is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the \strips\ action model $\Xi$}. A solution plan starts with a {\em prefix} that edits $\Xi$.
\item {\bf Validates the edited model $\Xi$ in $\mathcal{O}$}. The solution plan continues with a postfix that validates the edited model on the given observations $\mathcal{O}$.
\end{enumerate}

>>>>

Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016hierarchical,segovia2017generating}, we present a novel planning compilation approach for learning \strips\ action models. The output of the compilation is a classical planning task for which a solution plan is a sequence of actions that determine the precondition/effects of the model operator schemes (the learned model). \textbf{Learning action models as planning} leverages off-the-shelf planners and opens up a way towards the \emph{bootstrapping} of planning action models, enabling a planner to gradually learn and update its action model. The practicality of the compilation approach allow us to report results over a wide range of 15 IPC planning domains.

@@

@@

\subsection{Classical planning with conditional effects}
We use $F$ to denote the set of {\em fluents} (propositional variables) describing a state. A {\em literal} $l$ is a valuation of a fluent $f\in F$, i.e. either~$l=f$ or $l=\neg f$. A set of literals $L$ represents a partial assignment of values to fluents (WLOG we assume that $L$ does not assign conflicting values to any fluent). We use $\mathcal{L}(F)$ to denote the set of all literal sets on $F$, i.e.~all partial assignments of values to fluents.

@@

\section{Learning \strips\ action models}
%Learning \strips\ action models from plans where every action in the plan is available, as well as its corresponding {\em pre-} and {\em post-states}, is straightforward. In this case, \strips\ operator schemes are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Preconditions are derived lifting the minimal set of literals that appears in all the corresponding pre-states.

This paper addresses the learning task that corresponds to observing an agent acting in the world but watching only the results of its plan executions, the actual executed actions are unobserved. This learning task is defined as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}}$:
\begin{itemize}
\item $\mathcal{M}$ is the set of empty operator schemes, wherein each $\xi\in\mathcal{M}$ is only composed of $head(\xi)$.
\item $\Psi$ is the set of predicates, that define the abstract state space of a given classical planning frame.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$.
\end{itemize}

A solution to $\Lambda$ is a set of operator schema $\mathcal{M}'$ compliant with the headers in $\mathcal{M}$, the predicates $\Psi$, and the given sequence of state observations $\mathcal{O}$. A planning compilation is a suitable approach for addressing $\Lambda$ because a solution must not only determine the \strips\ action model $\mathcal{M}'$ but also, the {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$, that explains $\mathcal{O}$. Figure~\ref{fig:lexample} shows a $\Lambda$ task for learning a \strips\ action model in the blocksworld from the five-state observations sequence that corresponds to inverting a 2-block tower.

Our approach for addressing a $\Lambda$ learning task is compiling it into a classical planning task $P_{\Lambda}$ with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:

\begin{enumerate}
\item {\bf Programs the \strips\ action model $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that, for each $\xi\in\mathcal{M}$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item {\bf Validates the \strips\ action model $\mathcal{M}'$ in $\mathcal{O}$}. The solution plan continues with a postfix that produces the given sequence of states $\tup{s_0,s_1,\ldots,s_{n}}$ using the programmed action model $\mathcal{M}'$.
\end{enumerate}

\begin{figure}
{\footnotesize\tt ;;;;;; Headers in $\mathcal{M}$}
\begin{footnotesize}
\begin{verbatim}
(pickup v1) (putdown v1)
(stack v1 v2} (unstack v1 v2)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;;;;; Predicates $\Psi$}
\begin{footnotesize}
\begin{verbatim}
(handempty) (holding ?o  - object)
(clear ?o - object) (ontable ?o - object)
(on ?o1 - object ?o2 - object)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;;;;; Observations $\mathcal{O}$}
\begin{footnotesize}
\begin{verbatim}
;;; observation #0
(clear block2) (on block2 block1)
(ontable block1) (handempty)

@@

Given a learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ built instantiating the predicates $\Psi$ with the objects appearing in the input observations $\mathcal{O}$.
\item Fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$, that represent the programmed action model. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative/positive effect in the schema $\xi\in \mathcal{M}'$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by the pair of fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to {\em True}.
\item The fluents $mode_{prog}$ and $mode_{val}$ indicating whether the operator schemes are programmed or validated and fluents $\{test_i\}_{1\leq i\leq n}$, indicating the observation where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. Our compilation assumes that initially operator schemes are programmed with every possible precondition, no negative effect and no positive effect. Therefore fluents $pre_f(\xi)$, for every $f\in F_v(\xi)$, hold also at the initial state.
\item $G_{\Lambda}=\bigcup_{1\leq i\leq n}\{test_i\}$, indicates that the programmed action model is validated in all the input observations.
\item $A_{\Lambda}$ comprises three kinds of actions:
\begin{enumerate}
\item Actions for {\em programming} operator schema $\xi\in\mathcal{M}$:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\mathcal{M}$.

@@

\item Actions for {\bf adding} a {\em negative} or {\em positive} effect $f\in F_v(\xi)$ to the action schema $\xi\in\mathcal{M}$.

@@

\item Actions for {\em applying} an already programmed operator schema $\xi\in\mathcal{M}$ bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Given that the operators headers are known, the variables $pars(\xi)$ are bound to the objects in $\omega$ appearing at the same position. 
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
&\cup \{\neg mode_{val}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\},\\
&\{\emptyset\}\rhd\{mode_{val}\}.
\end{align*}
\end{small}

\item Actions for {\em validating} an observation {\tt\small $1\leq i\leq n$}.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&s_i\cup\{test_j\}_{j\in 1\leq j<i}\\
&\cup\{\neg test_j\}_{j\in i\leq j\leq n}\cup \{mode_{val}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i,\neg mode_{val}\}.
\end{align*}
\end{small}
\end{enumerate}
\end{itemize}

@@

\begin{lemma}
Soundness. Any classical plan $\pi$ that solves $P_{\Lambda}$ induces an action model $\mathcal{M}'$ that solves the $\Lambda$ learning task.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
Once operator schemes $\mathcal{M}'$ are programmed, they can only be applied because of the $mode_{prog}$ fluent. In addition, $P_{\Lambda}$ is only solvable if fluents $\{test_i\}$, {\small $1\leq i\leq n$} hold at the last reached state. These goals can only be achieved executing an applicable sequence of programmed operator schemes that reaches every state $s_i\in\mathcal{O}$, starting from $s_0$ and following the sequence {\small $1\leq i\leq n$}. This means that the programmed action model $\mathcal{M}'$ complies with the provided observations $\mathcal{O}$ and hence, solves $\Lambda$.
\end{small}
\end{proof}

\begin{lemma}
Completeness. Any \strips\ action model $\mathcal{M}'$ that solves a $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}}$ learning task, is computable solving the corresponding classical planning task $P_{\Lambda}$.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
The compilation does not discard any possible \strips\ action schema definable within $F_v$ that satisfy the state trajectory constraints given by the $\mathcal{O}$ sequence. By definition, $F_v(\xi)\subseteq F_\Lambda$ fully captures the full set of elements that can appear in a \strips\ action schema $\xi\in\mathcal{M}$ given its header and the set of predicates $\Psi$.
\end{small}
\end{proof}

The size of $P_{\Lambda}$ depends on the number of given state observations, that is $|\mathcal{O}|$. The larger this number, the more $test_i$ fluents and $\mathsf{validate_{i}}$ actions.

\section{Evaluation of \strips\ action models}
We assess how well a \strips\ action model $\mathcal{M}$ explains a given sequence of observations $\mathcal{O}$ according to the amount of {\em edition} that is required by $\mathcal{M}$ to produce $\mathcal{O}$. %In the extreme, if $\mathcal{M}$ perfectly explains $\mathcal{O}$, no model {\em edition} is necessary.

\subsection{Edition of \strips\ action models}
Let us define first the \emph{operations} allowed to edit a given \strips\ action model. With the aim of keeping tractable the branching factor of the planning instance that results from our compilation, we only define two {\em edit operations}:
\begin{itemize}
\item {\em Deletion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ is removed from the operator schema $\xi\in\mathcal{M}$, $f\in F_v(\xi)$.
\item {\em Insertion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ is added to the operator schema $\xi\in\mathcal{M}$, $f\in F_v(\xi)$.
\end{itemize}

Now, we can formalize an edit distance that quantifies how dissimilar two given \strips\ action models are. The distance is symmetric and meets the {\em metric axioms} provided that the two {\em \strips\ edit operations} have the same positive cost.

\begin{definition}
Let $\mathcal{M}$ and $\mathcal{M}'$ be two \strips\ action models, both built from the same set of possible elements $F_v$. The {\bf edit distance}, denoted as $\delta(\mathcal{M},\mathcal{M}')$, is the minimum number of {\em edit operations} required to transform $\mathcal{M}$ into $\mathcal{M}'$.
\end{definition}

Since $F_v$ is a bound set, the maximum number of edits that can be introduced to a given action model defined within $F_v$ is bound as well. In more detail, for an operator schema $\xi\in\mathcal{M}$ the maximum number of edits that can be introduced to their precondition set is $|F_v(\xi)|$ while the max number of edits that can be introduced to the effects are two times $|F_v(\xi)|$.

\begin{definition}
The \textbf{maximum edit distance} of an \strips\ action model $\mathcal{M}$ built from the set of possible elements $F_v$ is $\delta(\mathcal{M},*)=\sum_{\xi\in\mathcal{M}} 3|F_v(\xi)|$ and sets an upperbound on the distance from $\mathcal{M}$ to any \strips\ model definable within $F_v$.
\end{definition}

@@

\begin{definition}
Given an action model $\mathcal{M}$ built from the set of possible elements $F_v$ and a sequence of observations $\mathcal{O}$. The {\bf observation edit distance}, denoted by  $\delta(\mathcal{M},\mathcal{O})$, is the minimal edit distance from $\mathcal{M}$ to any model $\mathcal{M}'$ such that: (1) $\mathcal{M}'$ is also definable within $F_v$ and (2), $\mathcal{M}'$ can produce a plan $\pi=\tup{a_1, \ldots, a_n}$ that induces $\mathcal{O}=\tup{s_0, s_1, \ldots, s_n}$; \[\delta(\mathcal{M},\mathcal{O})=\min_{\forall \mathcal{M}' \rightarrow \mathcal{O}} \delta(\mathcal{M},\mathcal{M}')\]
\end{definition}

Because its semantic nature, the {\em observation edit distance} is robust to evaluate learning episodes where actions are {\em reformulated} and still compliant with the learning inputs. For instance a learning episode can interchange the roles of operators with matching headers, or the roles of action parameters with matching types (e.g. the {\em blocksworld} operator {\small\tt stack} can be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa).

Furthermore, the {\em observation edit distance} enables the recognition of \strips\ action models. The idea, taken from {\em plan recognition as planning}~\cite{ramirez2009plan}, is to map distances into likelihoods. The {\em observation edit distance} can be mapped into a likelihood with the following expression $P(\mathcal{O}|\mathcal{M})=1-\frac{\delta(\mathcal{M},\mathcal{O})}{\delta(\mathcal{M},*)}$. The larger the {\em observation edit distance} the lower this likelihood.

%Given a set of possible \strips\ models and a sequence of state observations, the {\em recognition of \strips\ models} is the task of computing the model with the highest probability according to the cited observations. According to the Bayes rule, the probability of an hypothesis $\mathcal{H}$ given the observations $\mathcal{O}$ can be computed with $P(\mathcal{H}|\mathcal{O})=\frac{P(\mathcal{O}|\mathcal{H})P(\mathcal{H})}{P(\mathcal{O})}$. In our scenario, the hypotheses are about the possible \strips\ action models that can be built within a given set of predicates $\Psi$ and a given a set of operator headers (in other words, given the $F_v(\xi)$ sets). With this regard, $P(\mathcal{M}|\mathcal{O})$, the probability distribution of the possible \strips\ models (within the $F_v(\xi)$ sets) given an observation sequence $\mathcal{O}$ could be computed by:
%\begin{enumerate}
%\item Computing the {\em observation edit distance} $\delta(\mathcal{M},\mathcal{O})$ for every possible model $\mathcal{M}$.
%\item Applying the resulting distances to the above $P(\mathcal{O}|\mathcal{M})$ formula to map these distances into likelihoods
%\item Applying the Bayes rule to obtain the normalized posterior probabilities, these probabilities must sum 1.
%\end{enumerate}

\begin{figure}
{\tt\small
00 : (program\_add\_handempty\_stack)\\
01 : (program\_add\_clear\_stack\_var1)\\
02 : (apply\_unstack block2 block1)\\
03 : (validate\_1)\\
04 : (apply\_putdown block2)\\
05 : (validate\_2)\\
06 : (apply\_pickup block1)\\
07 : (validate\_3)\\
08 : (apply\_stack block1 block2)\\
09 : (validate\_4)\\     
}
 \caption{\small Plan for editing a given {\em blockswold} schema and validating it at the state observations shown in Figure~\ref{fig:lexample}.}
\label{fig:plan}
\end{figure}

\subsection{Computing the observation edit distance}
Our compilation is extensible to compute the {\em observations edit distance}. A solution to the planning task resulting from the extended compilation is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the \strips\ action model $\mathcal{M}$}. A solution plan starts with a {\em prefix} that edits the preconditions and effects of the action schemes in $\mathcal{M}$.
\item {\bf Validates the edited model $\mathcal{M}$ in $\mathcal{O}$}. The solution plan continues with a postfix that validates the edited model on the given observations $\mathcal{O}$.
\end{enumerate}
